---
title: "Testing"
author: "Alexandra Niessl"
date: "12/16/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Context
The function 'getSimulatedData' creates a single simulated data set for a single treatment arm.
The function is used as basis for simulating a large number of oncology clinical trials. The large number of simulation iterations is generated using the function 'replicate'. The input parameters of 'getSimulatedData' can vary for each treatment arm. 
That is, the simulation of a trial with nRep simulation iterations will be constructed as follows:

replicate(nRep, function(
  simulateTrt<-getSimulatedData(...)
  simulatePlac<-getSimulatedData(...)
  combine+coding treatment
  )
)

For most applications, the data sets are converted to a wide format (one row per individual).
The simulations are used to answer different planning questions for clinical trials with endpoints overall survival and progression-free survival.


## Objective
The function 'getSimulatedData' simulates data from an illness-death model with one row per transition and subject. It allows for random dropouts, staggered study entry, different distributions of event times (exponential, piece-wise exponential, Weibull)
Input parameters:
N:    Number of patients
transition: specifies the three transition hazards of the illness-death model
            see class 'transition'
            default: exponential distributed event times with h01=h02=h12=1
dropout:   specifies dropout probability (dropout\$rate) 
           within a time range dropout\$time
           default: no random  dropout/censoring
accrual:   specifies the recruiting/staggered study entry
           Either accrual intensity (number of patients recruited per time unit) 
           or accrual time  could be specified, see function 'addStaggeredEntry'
           default: no staggered study entry

Output:
simulated data set with the following columns:
id: patient id
from: starting state of transition
to: final state of transition
entry: entry time of 'from' on individual time  scale
exit: exit time of 'from' and entry time of 'to' on individual time scale
entryAct: entry time on study time scale
exitAct:  exit time on study time scale
censAct:  censoring time on study time scale  (required?)


Illness-death multistate data are simulated using the following algorithm:
1. simulate waiting times in the initial state with all-cause hazard h01(t)+h02(t)
2. Run a binomial experiment for a simulated waiting-time, say $t_0$, which decides with probability
h01 ($t_0$) /(h01($t_0$)+h02($t_0$)) on progression 
3. If the binomial experiment decides on death, step 4 follows. If the binomial experiment decides on
intermediate state progression, the waiting time in the progression state is simulated with hazard h12(t), t $\geq t_0$.
4. If censoring is present: Additionally, generate random exponential distributed censoring times (according to dropout parameters). Transition times of an individual occurring after the censoring time will be censored.
5.Staggered study entry times are added at the end - see helper function 'addStaggeredEntry'.


##Assertions
```{r}

library(checkmate)
##additional functions are needed to check input parameters of our function
assert_positive_number <- function(x) {
  assert_number(x)
  assert_true(x > 0)
}

```
##New class Transition
```{r}
exponential_transition <- function(h01, h02, h12) {
  assert_positive_number(h01)
  assert_positive_number(h02)
  assert_positive_number(h12)
  structure(
    list(
      hazards =   list(h01 = h01, h02 = h02, h12 = h12),
      intervals = list(pw01 = 0, pw02 = 0, pw12 = 0),
      weibull_rates = list(p01 = 1, p02 = 1, p12 = 1),
      family='exponential'
    ),
    class = "TransitionParameters"
  )
}

test <- exponential_transition(1, 1, 1)
class(test)


print.TransitionParameters <- function(object) {
  cat("Transition parameters for simulations with", unlist(object$family),             "distributed survival times \n")
  cat("Transition Hazards: ", "h01=", as.character(unlist(object$hazards[1])), ", h02=", unlist(object$hazards[2]), ", h12=", unlist(object$hazards[3]), sep=" ")
  ##add intervals and weibull rates (only for family = weibull, piecewise exponential..)
}

test

piecewise_exponential <- function(h01,h02,h12,pw01,pw02,pw12) {
  ##add check h01, h02, h12 vectors of number >0, p01, p02,p12 ..
  ##check intervals starting with 0
  ##add check length interval= length hazard
  structure(
    list(
      hazards =   list(h01 = h01, h02 = h02, h12 = h12),
      intervals = list(pw01 = pw01, pw02 = pw02, pw12 = pw12),
      weibull_rates = list(p01 = 1, p02 = 1, p12 = 1),
      family='piecewise exponential'
    ),
    class = "TransitionParameters"
  )
}

test_piec<-piecewise_exponential(c(1,1,1),c(1.5,0.5,1), c(1,1,1), c(0,3,5),c(0,6,7),c(0,8,9))
test_piec


weibull_transition <- function(h01,h02,h12,p01,p02,p12){
  assert_positive_number(h01)
  assert_positive_number(h02)
  assert_positive_number(h12)
  assert_positive_number(p01)
  assert_positive_number(p02)
  assert_positive_number(p12)
  structure(
  list(
    hazards = list(h01 = h01, h02 = h02, h12 = h12),
    weibull_rates = list(p01 = p01, p02 = p02, p12=p12),
    intervals = list(pw01 = 0,pw02 = 0,pw12 = 0),
    family="Weibull"
  ),
    class = "TransitionParameters"
  )
}
test_weibull <- weibull_transition(1, 1, 1,1.2,1.3,0.5)
class(test_weibull)
test_weibull 
```

## Prototype getSimulatedData

```{r prototype}

#' Title
#'
#' @param N (`int`)\cr number of patients.
#' @param transition (`TransitionParameters`)\cr transition parameters comprising 
#'   `hazards`, corresponding `intervals` and `weibull_rates`, see ...
#' @param dropout 
#' @param accrual 
#' 
#'
#' @return
#' @export
#'
#' @examples



getSimulatedData<-function(N,
                           transition = exponential_transition(h01 = 1, h02 = 1, h12 = 1),
                           dropout = list(rate = 0, time = 12),
                           accrual = list(param="time", value=0)) {
  ##check input parameters
  assert_int(N, lower = 1L)
  assert_class(transition, "TransitionParameters")
  assert(check_number(dropout$rate,lower=0,upper=1),
         check_number(dropout$time), 
         check_true(dropout$time > 0,),
         combine="and",.var.name="dropout")
  assert(check_choice(accrual$param, c("time","intensity")),
         check_number(accrual$value, lower=0), 
         combine="and",.var.name="accrual")
  
  #initialize transition hazards, Weibull rates and intervals
  h01<-unlist(transition$hazard['h01'])
  h02<-unlist(transition$hazard['h02'])
  h12<-unlist(transition$hazard['h12'])
  
  p01<-unlist(transition$weibull_rates['p01'])
  p02<-unlist(transition$weibull_rates['p02'])
  p12<-unlist(transition$weibull_rates['p12'])
  
  pw01<-unlist(transition$intervals['pw01'])
  pw02<-unlist(transition$intervals['pw02'])
  pw12<-unlist(transition$intervals['pw12'])
  
  #get rate parameter for exponential distributed censoring times 
  #if rate=0, no censoring is applied
  censRate<- ifelse(dropout$rate!=0,-log(1-dropout$rate)/dropout$time,0)
  #censoring times for all individuals, infinity if no censoring is applied
  censTime<-ifelse(rep(dropout$rate,N)!=0,rexp(N, censRate),rep(Inf,N))
 
  ###all individuals start in the initial state
  entry<-rep(0,N)
  from<-rep(0,N)
  ###Waiting time in the initial state 0 
  U<-runif(N)
  ##for exponential or Weibull distributed survival times
  if(transition$family %in% c("exponential", "Weibull")){
    ## for distribution of waiting time in the initial state all-cause hazard (h01+h02) is  needed
    wait_time<- getWaitTimeSum(N,U,h01,h02,p01,p02,max.int=10000,entry)  
    ###binomial experiment decides on death or progression
    numerator<-p01*h01^p01*wait_time^(p01-1)
    denumerator<-numerator+p02*h02^p02*wait_time^(p02-1)
 
   ###for piecewise exponential distributed survival times
   }else if(transition$family =="piecewise exponential"){
    wait_time<- getPCWDistr(N,U, getSumPCW(h01, h02,pw01, pw02), 
                    unique(sort(c(pw01,pw02))),entry)
    ###binomial experiment decides on death or progression
    numerator<-getPCWHazard(h01,pw01,wait_time)
    denumerator<-getPCWHazard(h01,pw01,wait_time)+getPCWHazard(h02,pw02,wait_time)
   }
  
  to_prob <- rbinom(N,1, numerator/denumerator)
  to<- ifelse(to_prob==0,2,1)
  exit<-wait_time
  
  ##add censoring
  to<-ifelse(censTime< wait_time, "cens", to)  
  exit<-pmin(censTime,wait_time)
  
  simData<-data.frame(id=1:N, from=from, to=to, entry=entry, exit=exit,
                      censTime=censTime,  stringsAsFactors = FALSE)
  ###1->2 transition
  id1<-simData$id[simData$to==1]
  if(length(id1)!=0){
    N1<-length(id1)
    U1<-runif(N1)
    to1<-rep(2,N1)
    from1<-rep(1,N1)
    entry1<-simData$exit[simData$id %in% id1]
    
    ##Wait time in state 1
    ##exponential 
    if(transition$family=="exponential"){
      wait_time1<- (-log(1-U1))/(h12)
    ##Weibull
    }else if(transition$family=="Weibull"){
      wait_time1<- getWaitTimeSum(N1,U1,h12,h02=0,p12,p02=1,max.int=1000,entry=entry1)
    #piecewise
    }else if(transition$family=="piecewise exponential"){
      wait_time1<- getPCWDistr(N1,U1,haz=h12, pw=pw12,t_0=entry1)
    }
    exit1<-entry1+wait_time1
    ##add censoring
      censTime1<-censTime[id1]
      to1<-ifelse(censTime1< exit1, "cens", to1)  
      exit1<-pmin(censTime1,exit1)
    
    newRows<-data.frame(id=id1, from=from1, to=to1, entry=entry1,exit=exit1,
                        censTime=censTime1,stringsAsFactors = FALSE)
    simData<-rbind(simData,newRows)
  }  
  
  simData<- simData[order(simData$id),]
  ##add staggered study entry, i.e. study entry at calendar time
  simData<-addStaggeredEntry(simData=simData, N=N,
                             accrualParam=accrual$param, accrualValue=accrual$value)
  
  return(simData)
} 

```

## Helper functions
getsimulateData uses the following helper functions:

getPCWHazard: for piecewise-constant hazards: function to get hazard value at time x
getSumPCW: calculates the sum of two piece-wise constant hazards
getPCWDistr: generates event times with a distribution resulting 
            from specified piece-wise constant hazards 
getWaitTimeSum: get event times following the distribution of a sum of
                Weibull distributed event-times 
addStaggeredEntry: add staggered study entries to study. This
                   implies that times on individual scale and 
                   study scale differ

```{r}

addStaggeredEntry<-function(simData,N, accrualParam, accrualValue){
  ##get accrual times in calendar time per individual.
  ##if no staggered study entry, all individuals have the same entry time 0
   if(accrualValue!=0){
     if(accrualParam == "time"){
        entry_act<-runif(N,0,accrualValue)
   }
     if(accrualParam == "intensity"){
        accrualTime<-N/accrualValue 
        entry_act<-runif(N,0,accrualTime)
     } 
   }else if(accrualValue==0){
     entry_act=rep(0,N)
   }
   entryAct<-cbind(id=1:N,entry_act)
  # combine simulated data with actual entry time
  # and generate other actual time (actual entry time+ individual time)
    simData<-merge(simData, entryAct)    
    simData$entryAct<-simData$entry+simData$entry_act
    simData$exitAct<-simData$exit+simData$entry_act
    simData$censAct<-simData$censTime+simData$entry_act
   #delete temporary helper columns
    simData[,c("entry_act")]<-list(NULL)
    simData[,c("censTime")]<-list(NULL)
   return(simData)
}
 

### Helper functions for the simulation of MSM data with piecewise constant hazard
### or Weibull distributed event times

##for piecewise-constant hazards: function to get hazard value at time x
getPCWHazard<-function(haz,pw,x){
  hazVal<-sapply(x,function(jj){
    y<-NULL 
    #find interval and correponding hazard value for time x[jj]
    for(ii in 1:length(haz)){
      if(jj>=pw[ii])
        y<-haz[ii]
    }
    hazVal<-y
    return(hazVal)
  }) 
  return(hazVal)
}

#function to calculate sum of two piece-wise constant hazards
getSumPCW<-function(haz1,haz2, pw1,pw2){ 
  #get all cutpoints for the intervals
  cuts_sum<- unique(sort(c(pw1,pw2)))
  haz_sum<-NULL
  ##get sum of hazards for all intervals
  for(i in 1:(length(cuts_sum))){
    haz_sum[i]<-getPCWHazard(haz1,pw1,cuts_sum[i])+getPCWHazard(haz2,pw2,cuts_sum[i])
  }
  return(haz_sum)
}

###function to get event times with a distribution resulting from piece-wise constant hazards
getPCWDistr<-function(N,U,haz,pw,t_0){ 
  t1 <- rep(NA,N) #event times
  n2<-length(haz) 
   for(kk in 1:N){
    ##shift if t_0 !=0
    cuts_temp<-c(t_0[kk],pw[pw>t_0[kk]])
    cuts_temp<-cuts_temp-t_0[kk]
    #number of cutpoints of intervals
    n<-length(cuts_temp)
    ##number of hazards to remove for shift (t_0> as respective hazard intervals)
    remov<-n2-n
    haz_temp<-haz[(remov+1):n2] 
    LogU<-log(1-U[kk])
   
    if(n!=1){
      ##determine sum of alpha*time-interval for all i
      dt<-cuts_temp[2:n]-cuts_temp[1:(n-1)]
      
      #helping matrix
      tempMatrix<-matrix(0, nrow=n,ncol=n-1)
      tempMatrix[lower.tri(tempMatrix)]<-1
      
      sumA<--as.vector(tempMatrix%*%(haz_temp[1:(n-1)]*dt))
      
      # find the appropriate time interval
      for (i in 1:n) {
        if(i !=n){
          t1[kk] <- ifelse(sumA[i]>= LogU&  LogU>sumA[i+1],
                           cuts_temp[i] + (sumA[i] -LogU)/haz_temp[i],
                           t1[kk])
        }else{
           t1[kk] <- ifelse(LogU <= sumA[i], 
                           cuts_temp[i] + (sumA[i] -LogU)/haz_temp[i], 
                           t1[kk])
        }
      }
    }else if(n==1){
      ##only one interval ->exponential distribution
      t1[kk]<- -LogU/haz_temp
    } 
  }
  return(t1)
}

###Function to get event times following the distribution of a sum of Weibull distributed event-times

getWaitTimeSum<-function(N,U,h01,h02,p01,p02,max.int,entry){
  ##exponential distributed survival times
  if(p01 == 1 & p02 ==1){
    return(-log(1-U)/(h01+h02))
  }else{
  #Weibull distributed survival times
  temp<-function(x,y,t0){
        return((h01*(x+t0))^p01-(h01*t0)^p01-(h02*t0)^p02+(h02*(x+t0))^p02+y)}
  stime<-NULL
  i<-1
    while (length(stime) < N){
      u<-U[i]
      t0<-entry[i]  
      if(temp(0,log(1-u),t0)*temp(max.int, log(1-u),t0)<0){
        res<-uniroot(temp,c(0,max.int), tol=10^-16,y=log(1-u),t0=t0)
        stime[i]<-res$root
        i<-i+1
        if(res$root==0) {cat("Check tol argument of uniroot")}
      }
      else cat("Waiting time in the inital state -uniroot: Values at endpoints not of opposite signs. \n")
    }
  return(stime)
  }
}
```


## Try it out

```{r}
set.seed(12345)
# 1-2 examples
testExp<-getSimulatedData(20,
                 transition = exponential_transition(h01 = 1, h02 = 1.5, h12 = 1),
                 dropout = list(rate = 0.5, time = 1),
                 accrual = list(param="time",value=5)) 

testWeibull<-getSimulatedData(20,
                 transition = weibull_transition(h01 = 1, h02 = 1.5, h12 = 1, p01=1.3, p02=1.2, p12=0.5),
                 dropout = list(rate = 0.5, time = 1),
                 accrual = list(param="time", value = 0)) 

testPiecExp<-getSimulatedData(20,
                 transition = piecewise_exponential(h01 = c(1,1.2,1.3), h02 = c(1.5,1.4,1.2), h12 = c(1,0.6), pw01=c(0,4,5), pw02=c(0,1,2), pw12=c(0,3)),
                 dropout = list(rate = 0, time = 12),
                 accrual = list(param="intensity", value = 7)) 
```

