---
title: "test for getSimulatedData"
author: "Alexandra Niessl"
date: "4/8/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Objective

Create a test that is able to check whether the transition hazards of the MSM lead to the specified distribution of
survival times.
Idea: Compare true the Nelson-Aalen estimator (can be derived from specified transitions) with the mean Nelson-Aalen over 1000 simulations. Or: empirical distribution of survivla times (without censoring)?
Use `skip_on_ci()` as the long running test should be skiped on GitHub.



##Compare TRUE Nelson-Aalen with simulated

```{r}
library(mvna)

nRep <- 1000
N <- 200

tra <- matrix(ncol = 3, nrow = 3, FALSE)
tra[1, 2:3] <- TRUE
tra[2, c(3)] <- TRUE


## averaged Nelson-Aalen
estimateNA <- function(simTrials, tr.choice, times) {
  EstimatedNA <- lapply(seq_along(simTrials), function(j) {
    na <- NULL
    na <- mvna(simTrials[[j]], c("0", "1", "2"), tra, "cens")
    na_predict <- predict(na, times,
      tr.choice = c(tr.choice),
      level = 0.95, var.type = c("aalen"), ci.fun = c("log")
    )[[tr.choice]][, "na"]
    return(na_predict)
  })
  EstimatedNAMean <- rowMeans(matrix(unlist(EstimatedNA), ncol = length(EstimatedNA)))
  return(EstimatedNAMean)
}



### constant transition hazards
startTime <- Sys.time()
simTrials1 <- replicate(nRep, getSimulatedData(N,
  transition = exponential_transition(h01 = 0.3, h02 = 0.4, h12 = 0.4),
  dropout = list(rate = 0, time = 12),
  accrual = list(param = "time", value = 0)
), simplify = FALSE)

times <- seq(0, 20, 0.1)
Const12 <- estimateNA(simTrials1, "1 2", times)
Const01 <- estimateNA(simTrials1, "0 1", times)

# 1->2 transition
## true alpha
trueAlpha <- 0.4
trueNA <- trueAlpha * times
#### plot
plot(times, Const12, type = "l")
lines(times, trueNA, col = "red")


# 0->1 transition
## true alpha
trueAlpha <- 0.3
trueNA <- trueAlpha * times
#### plot
plot(times, Const01, type = "l")
lines(times, trueNA, col = "red")
endTime <- Sys.time()

# prints recorded time
print(endTime - startTime)
```


Add for Weibull and piecewise transition hazards. True Nelson-Aalen estimators:

```{r}
# Weibull
trueNA <- (trueAlpha * times)^p

### PW constant
truePW <- function(alpha, pw, times) {
  if (times <= pw[2]) {
    return(alpha[1] * (times - pw[1]))
  } else if (times <= pw[3]) {
    return(alpha[1] * (pw[2] - pw[1]) + alpha[2] * (times - pw[2]))
  } else {
    return(alpha[1] * (pw[2] - pw[1]) + alpha[2] * (pw[3] - pw[2]) + alpha[3] * (times - pw[3]))
  }
}
```
